---
title: "01_exploratory_data_analysis"
author: "Nguyen Trung Nam"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

LOAD LIBRARY & USER-DEFINED FUNCTIONS
```{r Library, echo=FALSE}
library(data.table)
library(tidyverse)
library(raster)
library(rgdal)
library(sf)
library(readr)
library(lubridate)
library(zoo)
library(forcats)
library(corrr)
library(cowplot)
library(fitdistrplus)
library(xts)
library(qmap)
library(car)
source("d:/Repository/R_Postgres/R/F-R_conect_DB.R")
source("d:/Repository/Mike11model/R/f-TL_DBSCL_model.R")
source("d:/Repository/R_Postgres/R/Thongso_chung_vebieudo.R")
source("d:/Repository/R_Postgres/R/F-summary.R")
source("d:/Repository/R_Postgres/R/F-fixExten_areaClassify_raster.R")
# Phần chỉnh thông số chung - size & font-----
legsize     = 26  # Kích thước chữ legend
axisize     = 24  # Kích thước chữ của trục
axtilesize  = 22  # Kích thước chữ chú dẫn trục axis
tilesize    = 28  # Kích thước tên biểu đồ
legspace    = 1 # Khoảng cách các thành phần trong legend
fcetsize    = 22  # Kích thước chữ trong facet title
mary        = 2  # Khoảng cách axis title và axis trục y
marx        = 2  # Khoảng cách axis title và axis trục x
antext      =  5  # Kích thước size text của annotation
ansym       =  6  # Kích thước segment của legend
```

# 01:PREPARE GLOBAL REANALYZED PRECIPITATION
Aims:
-> To extract reanalyzed rainfall which provided by NOAA at given stations
-> To pre-assess observed precipitation data of these stations
```{r}
# extract the precipitation data-----
path_nc      <- "d:/Repository/R_Postgres/rawdata/ncdf/"
path_shp     <- "d:/Trung Nam/SIWRP/Nam/Dubaolu_dautieng_2023/04_Gis/data/"
sta_loca     <- st_read(paste0(path_shp,"trammua_lvdautieng_latlon.shp")) %>%
                as.data.frame()
sta_list     <- sta_loca %>% mutate(no = row_number()) %>% dplyr::select(no,sta) %>%
                mutate(station_id = sta, sta = NULL)

sta_loca     <- sta_loca %>% mutate(longitute  = unlist(map(sta_loca$geometry,1)),
                                    latitute   = unlist(map(sta_loca$geometry,2)))
sta_loca     <- sta_loca %>% dplyr::select(longitute,latitute)
## for loop to extract rainfall 
dt    <- as.numeric()
files <- list.files(paste0(path_nc),pattern = ".nc")
for (i in 1:length(files))
{
  ncd_r <- brick(paste0(path_nc,files[i]))
  df    <- raster::extract(ncd_r,sta_loca,method="bilinear")
  df    <- as.data.frame(df)
  df    <- t(df) %>% as.data.frame()
  df    <- df %>% tibble::rownames_to_column()
  df    <- gather(df,no,value,-rowname)
  df    <- separate(df,rowname, into = c('y','m','d','xb1','xb2','xb3','xb4'), sep = '\\.')
  df    <- df %>% dplyr::select(y,m,d,no,value)
  df    <- df %>% mutate(y  = gsub(pattern = "X",'',y),
                         no = gsub(pattern = "V",'',no),
                         across(c(y,m,d,no), as.numeric))
  dt    <- rbind(df,dt)
}
## join to get station name
df_th <- left_join(dt, sta_list, by = 'no') %>% fdate()
df_th <- df_th %>% dplyr::select(y,m,d,date,station_id,value)
df_th <- spread(df_th,station_id,value)
write.csv(df_th, file = 'Noaa_rain_trammuaDT.csv',sep = ',',row.names = F)

# quickly assess statistical basic----
precip_data <- read_csv("d:/Repository/ML_SelfStudy_FldForecast/Noaa_rain_trammuaDT.csv")
precip_data <- gather(precip_data,station_id,rainfall,-d,-m,-y,-date)

# Aggregate daily data into annual total rainfall
annual_rainfall <- precip_data %>%
  group_by(y,station_id) %>%
  summarise(total_annual_rainfall = sum(rainfall, na.rm = TRUE)) %>% ungroup()
# Plot annual rainfall boxplot
ggplot(annual_rainfall, aes(x = station_id, y = total_annual_rainfall)) +
  geom_boxplot() +
  labs(title = "Annual Rainfall Boxplot", x = "Year", y = "Total Annual Rainfall (mm)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Aggregate daily data into monthly total rainfall (across all years)
monthly_rainfall <- precip_data %>%
  group_by(m,station_id) %>%
  summarise(total_monthly_rainfall = sum(rainfall, na.rm = TRUE))

# Plot monthly rainfall boxplot
ggplot(monthly_rainfall, aes(x = factor(m, labels = c('Jan', 'Feb', 'Mar', 'Apr',
                                                          'May', 'Jun', 'Jul', 'Aug',
                                                          'Sep', 'Oct', 'Nov', 'Dec')), 
                             y = total_monthly_rainfall)) +
  geom_boxplot() +
  labs(title = "Monthly Rainfall Boxplot", x = "Month", y = "Total Monthly Rainfall (mm)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 02:BIAS CORRECTION USING QMAP PACKAGE
Aims:
-> To familiarize my self with functions in the Qmap package
-> Sample data: 
                + Selection stations: Tay Ninh, Loc Ninh
                + Calibration period: 1991 - 2000
                + Validation period: 2001 - 2021
```{r}
# data preparation -----
obsprecip      <- read_excel("d:/Repository/ML_SelfStudy_FldForecast/rawdata/obs_precip_flow_dtbasin.xlsx",
                             sheet = "rain")
obsprecip      <- obsprecip %>% mutate(across(-date,as.numeric))
## retrieve obs rainfall data for calibration
obsprecip      <- obsprecip %>% dplyr::select(date,tayninh,locninh) %>% 
                       mutate(yr = year(date)) 
obspre_train   <- obsprecip %>% mutate(y = year(date),m = month(date), d = day(date)) %>%
                    fdate() %>% filter(yr > 1990, yr < 2001) %>% dplyr::select(date,tayninh,locninh) 
obspre_test    <- obsprecip %>% mutate(y = year(date),m = month(date), d = day(date)) %>%
                    fdate() %>% filter(yr > 2001, yr < 2019) %>% dplyr::select(date,tayninh,locninh) 
write.csv(obspre_test,file  = "xb_obs_test.csv", sep = ",", row.names = F)
## convert to an xts object
obspre_train   <- xts(x = obspre_train[, -which(names(obspre_train) == "date")], order.by = obspre_train$date)
obspre_test    <- xts(x = obspre_test[, -which(names(obspre_test) == "date")], order.by = obspre_test$date)

# modeled precipitation
simprecip      <- read_csv("d:/Repository/ML_SelfStudy_FldForecast/rawdata/precip_dtbasin_noaa.csv")
## retrieve noaa rainfall data for calibration
simpre_train   <- simprecip %>% dplyr::select(y,m,d,tayninh,locninh) %>%
                  filter(y > 1990, y < 2001) %>% fdate() 
simpre_test    <- simprecip %>% dplyr::select(y,m,d,tayninh,locninh) %>%
                  filter(y > 2001,y < 2019) %>% fdate() 
write.csv(simpre_test,file  = "xb_sim_test.csv", sep = ",", row.names = F)
## convert to an xts object
simpre_train   <- xts(x = simpre_train[, -which(names(simpre_train) == "date")], order.by = simpre_train$date)
simpre_test    <- xts(x = simpre_test[, -which(names(simpre_test) == "date")], order.by = simpre_test$date)

## retrieve vectors
obs_train     <- as.vector(obspre_train$tayninh) 
obs_train     <- obs_train[obs_train > 0] %>% na.omit()

sim_train     <- as.vector(simpre_train$tayninh) 
sim_train     <- sim_train[sim_train > 0] %>% na.omit()
## extracting test data
obs_test     <- as.vector(obspre_test$tayninh)%>% na.omit()
sim_test     <- as.vector(simpre_test$tayninh)%>% na.omit()

# testing parametric transformation-------
fit_ptf   <- fitQmapPTF(obs_train,sim_train, transfun = "expasympt.x0",
                       cost = "MAE")
pre_fixed <- doQmapPTF(sim_test,fit_ptf)

## assessing changes in monthly or yearly precipitation amount
dt_test    <- data.frame(sim   = sim_test,
                         fixe  = pre_fixed,
                         obs   = obs_test,
                         y     = simpre_test$y,
                         m     = simpre_test$m,
                         d     = simpre_test$d)
dt_y      <-  gather(dt_test,type_id,value, - y, -m, -d)
dt_y      <-  dt_y %>% group_by(y,type_id) %>% mutate(value = sum(value)) %>% ungroup() %>%
            dplyr::select(y,type_id,value) %>% distinct()
dt_y      <- spread(dt_y,type_id, value)
write.csv(dt_y, file = "xb_parametric_corrected.csv", sep = ",", row.names = F)


# model performance
## Root Mean Squared Error (RMSE)
rmse <- sqrt(mean((pre_fixed - obs_test)^2))
cat("RMSE_corrected:", rmse, "\n")
rmse <- sqrt(mean((sim_test - obs_test)^2))
cat("RMSE_none:", rmse, "\n")
# Mean Absolute Error (MAE)
mae <- mean(abs(pre_fixed - obs_test))
cat("MAE_corrected:", mae, "\n")
mae <- mean(abs(sim_test - obs_test))
cat("MAE_none:", mae, "\n")

# Pearson Correlation Coefficient (R)
correlation <- cor(pre_fixed, obs_test)
cat("R_corrected:", correlation, "\n")
correlation <- cor(sim_test, obs_test)
cat("R_none:", correlation, "\n")

# Visual comparison
plot(obs_test, pre_fixed, main="Observed vs Corrected Simulated Precipitation",
     xlab="Observed Precipitation", ylab="Corrected Simulated Precipitation",
     pch=19, col="blue")
abline(0, 1, col="red") # Line of perfect agreement

plot(obs_test, sim_test, main="Observed vs Simulated Precipitation",
     xlab="Observed Precipitation", ylab="Simulated Precipitation",
     pch=19, col="blue")
abline(0, 1, col="red") # Line of perfect agreement


# testing Quantile mapping using distribution derived transformations-------
fit_dist   <- fitQmapDIST(obs_train,sim_train)
pre_fixed <- doQmapPTF(sim_test,fit_dist)

# model performance
## Root Mean Squared Error (RMSE)
rmse <- sqrt(mean((pre_fixed - obs_test)^2))
cat("RMSE_corrected:", rmse, "\n")
rmse <- sqrt(mean((sim_test - obs_test)^2))
cat("RMSE_none:", rmse, "\n")
# Mean Absolute Error (MAE)
mae <- mean(abs(pre_fixed - obs_test))
cat("MAE_corrected:", mae, "\n")
mae <- mean(abs(sim_test - obs_test))
cat("MAE_none:", mae, "\n")

# Pearson Correlation Coefficient (R)
correlation <- cor(pre_fixed, obs_test)
cat("R_corrected:", correlation, "\n")
correlation <- cor(sim_test, obs_test)
cat("R_none:", correlation, "\n")

# Visual comparison
plot(obs_test, pre_fixed, main="Observed vs Corrected Simulated Precipitation",
     xlab="Observed Precipitation", ylab="Corrected Simulated Precipitation",
     pch=19, col="blue")
abline(0, 1, col="red") # Line of perfect agreement

plot(obs_test, sim_test, main="Observed vs Simulated Precipitation",
     xlab="Observed Precipitation", ylab="Simulated Precipitation",
     pch=19, col="blue")
abline(0, 1, col="red") # Line of perfect agreement

# testing smoothing splines-------
fit_dist   <- fitQmapSSPLIN(obs_train,sim_train)
pre_fixed  <- doQmapSSPLIN(sim_test,fit_dist)

## assessing changes in monthly or yearly precipitation amount
dt_test    <- data.frame(sim   = sim_test,
                         fixe  = pre_fixed,
                         obs   = obs_test,
                         y     = simpre_test$y,
                         m     = simpre_test$m,
                         d     = simpre_test$d)
dt_y      <-  gather(dt_test,type_id,value, - y, -m, -d)
dt_y      <-  dt_y %>% group_by(y,type_id) %>% mutate(value = sum(value)) %>% ungroup() %>%
            dplyr::select(y,type_id,value) %>% distinct()
dt_y      <- spread(dt_y,type_id, value)
write.csv(dt_y, file = "xb_splines_corrected.csv", sep = ",", row.names = F)


# model performance
## Root Mean Squared Error (RMSE)
rmse <- sqrt(mean((pre_fixed - obs_test)^2))
cat("RMSE_corrected:", rmse, "\n")
rmse <- sqrt(mean((sim_test - obs_test)^2))
cat("RMSE_none:", rmse, "\n")
# Mean Absolute Error (MAE)
mae <- mean(abs(pre_fixed - obs_test))
cat("MAE_corrected:", mae, "\n")
mae <- mean(abs(sim_test - obs_test))
cat("MAE_none:", mae, "\n")

# Pearson Correlation Coefficient (R)
correlation <- cor(pre_fixed, obs_test)
cat("R_corrected:", correlation, "\n")
correlation <- cor(sim_test, obs_test)
cat("R_none:", correlation, "\n")

# Visual comparison
plot(obs_test, pre_fixed, main="Observed vs Corrected Simulated Precipitation",
     xlab="Observed Precipitation", ylab="Corrected Simulated Precipitation",
     pch=19, col="blue")
abline(0, 1, col="red") # Line of perfect agreement

plot(obs_test, sim_test, main="Observed vs Simulated Precipitation",
     xlab="Observed Precipitation", ylab="Simulated Precipitation",
     pch=19, col="blue")
abline(0, 1, col="red") # Line of perfect agreement


# testing non_parmater transformations-------
fit_dist   <- fitQmapQUANT(obs_train,sim_train)
pre_fixed  <- doQmapQUANT(sim_test,fit_dist)

## assessing changes in monthly or yearly precipitation amount
dt_test    <- data.frame(sim   = sim_test,
                         fixe  = pre_fixed,
                         obs   = obs_test,
                         y     = simpre_test$y,
                         m     = simpre_test$m,
                         d     = simpre_test$d)
dt_y      <-  gather(dt_test,type_id,value, - y, -m, -d)
dt_y      <-  dt_y %>% group_by(y,type_id) %>% mutate(value = sum(value)) %>% ungroup() %>%
            dplyr::select(y,type_id,value) %>% distinct()
dt_y      <- spread(dt_y,type_id, value)
write.csv(dt_y, file = "xb_nonparameter_corrected.csv", sep = ",", row.names = F)



# model performance
## Root Mean Squared Error (RMSE)
rmse <- sqrt(mean((pre_fixed - obs_test)^2))
cat("RMSE_corrected:", rmse, "\n")
rmse <- sqrt(mean((sim_test - obs_test)^2))
cat("RMSE_none:", rmse, "\n")
# Mean Absolute Error (MAE)
mae <- mean(abs(pre_fixed - obs_test))
cat("MAE_corrected:", mae, "\n")
mae <- mean(abs(sim_test - obs_test))
cat("MAE_none:", mae, "\n")

# Pearson Correlation Coefficient (R)
correlation <- cor(pre_fixed, obs_test)
cat("R_corrected:", correlation, "\n")
correlation <- cor(sim_test, obs_test)
cat("R_none:", correlation, "\n")

# Visual comparison
plot(obs_test, pre_fixed, main="Observed vs Corrected Simulated Precipitation",
     xlab="Observed Precipitation", ylab="Corrected Simulated Precipitation",
     pch=19, col="blue")
abline(0, 1, col="red") # Line of perfect agreement

plot(obs_test, sim_test, main="Observed vs Simulated Precipitation",
     xlab="Observed Precipitation", ylab="Simulated Precipitation",
     pch=19, col="blue")
abline(0, 1, col="red") # Line of perfect agreement




